---
title: "texas_blackout"
format: html
editor_options: 
  chunk_output_type: inline
---

## Setup

```{r}
#| message: false
#| warning: false

# Load required packages
library(tidyverse)
library(sf)
library(here)
library(terra)
library(stars)
library(tmap)
library(kableExtra)
library(basemaps)
```

## Data Loading and Preparation

```{r}
#| message: false
#| warning: false
# Read in the data

# Use rast() to read the night lights raster data tiles

VIIRS_07a <- rast(here("data/VNP46A1/VNP46A1.A2021038.h08v05.001.2021039064328.tif"))

VIIRS_07b <- rast(here("data/VNP46A1/VNP46A1.A2021038.h08v06.001.2021039064329.tif"))

VIIRS_16a <- rast(here("data/VNP46A1/VNP46A1.A2021047.h08v05.001.2021048091106.tif"))

VIIRS_16b <- rast(here("data/VNP46A1/VNP46A1.A2021047.h08v06.001.2021048091105.tif"))

# Use st_read() to read the gdb and gpkg files and query to extract only motorways and residential buildings

highways <- st_read(here("data/gis_osm_roads_free_1.gpkg/gis_osm_roads_free_1.gpkg"), 
                 query = "SELECT * FROM gis_osm_roads_free_1 WHERE fclass='motorway'", quiet = TRUE)

houses <- st_read(here("data/gis_osm_buildings_a_free_1.gpkg/gis_osm_buildings_a_free_1.gpkg"), 
                  query = "SELECT *
FROM gis_osm_buildings_a_free_1
WHERE (type IS NULL AND name IS NULL)
OR type in ('residential', 'apartments', 'house', 'static_caravan', 'detached')", quiet = TRUE)

socio <- st_read(here("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb/ACS_2019_5YR_TRACT_48_TEXAS.gdb/"), quiet = TRUE)
```

```{r}
# Make a data.frame containing the crs of the loaded datasets
crs_summary <- data.frame(
  Dataset = c("VIIRS tiles", "Highways", "Houses", "Socioeconomic"),
  EPSG = c(st_crs(VIIRS_07a)$epsg, 
           st_crs(highways)$epsg, 
           st_crs(houses)$epsg,
           st_crs(socio)$epsg))

# Display the crs
kable(crs_summary, 
      caption = "Coordinate Reference Systems of Input Data") %>% 
    kable_styling(bootstrap_options = c("striped", "hover"))
```

The socioeconomic data does not have a crs in its current form so we will have to extract the geometry and combine it with the relevant attributes

```{r}
#| message: false
#| warning: false

# Get the layers information
layers_info <- st_layers(here("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb/ACS_2019_5YR_TRACT_48_TEXAS.gdb/"))

# Display as scrollable table with kableExtra
kable(data.frame(
  Layer_Number = 1:length(layers_info$name),
  Layer_Name = layers_info$name,
  Geometry_Type = unlist(layers_info$geomtype)  
), 
      caption = "Available Layers in ACS Geodatabase") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                fixed_thead = TRUE) %>%
  scroll_box(height = "300px")
```

```{r}
#| message: false
#| warning: false

# Extract the geometry layer
socio_geom <- st_read(here("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb/ACS_2019_5YR_TRACT_48_TEXAS.gdb/"), 
                      layer = "ACS_2019_5YR_TRACT_48_TEXAS", quiet = TRUE)

# Extract the income layer
socio_income <- st_read(here("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb/ACS_2019_5YR_TRACT_48_TEXAS.gdb/"), 
                        layer = "X19_INCOME", quiet = TRUE)

# Rename the socio_income GEOID column to match the one in socio_geom
socio_income <- socio_income %>% 
  rename("GEOID_Data" = "GEOID")

# Join the income with the geometry by the GEOID_Data column
socio <- socio_geom %>% left_join(socio_income, by = "GEOID_Data") %>% 
  st_transform(st_crs(houses))

# Verify successful join
if(!("B19013e1" %in% names(socio))) {
  stop("Median household income (Estimate) column not found after join")
}
```

## Analysis

### Visualize Before and After
Map the difference in night light before and after the storms

```{r}
#| warning: false

# Merge the tiles for each date
VIIRS_07 <- merge(VIIRS_07a, VIIRS_07b)   # February 7, 2021
VIIRS_16 <- merge(VIIRS_16a, VIIRS_16b)   # February 16, 2021


# Crop, mask, and combine for plotting purposes
# 1. Define a Houston bounding box using the provided coordinates
HOUSTON_BBOX <- st_bbox(c(xmin = -96.5, ymin = 29, 
                          xmax = -94.5, ymax = 30.5), 
                        crs = st_crs(VIIRS_07))

# 2. Crop rasters using Houston bbox
crop_07 <- crop(VIIRS_07, HOUSTON_BBOX)
crop_16 <- crop(VIIRS_16, HOUSTON_BBOX)

# 3. Mask non-positive values and values above 1000
masked_07 <- crop_07
masked_07[(masked_07 > 1000) | (masked_07 <= 0)] <- NA

masked_16 <- crop_16
masked_16[(masked_16 > 1000)  | (masked_07 <= 0)] <- NA

# 4. Combine cropped and masked rasters
combined_crop <- c(masked_07, masked_16)

# Name the layers
names(combined_crop) <- c("February 7, 2021 (Before)", "February 16, 2021 (After)")

# Plot the before and after
tm_shape(combined_crop) +
  tm_raster(
    palette = "inferno",,
    title = "Night Light Intensity"
  ) +
  tm_facets(ncol = 2, free.scales = FALSE) +
  tm_layout(
    main.title = "Houston Night Lights Before and After Feb 2021 winter Storm Blackouts",
    main.title.size = 1.2,
    legend.outside = TRUE,
    legend.title.size = 0.8,
    legend.text.size = 0.6,
    legend.bg = FALSE,
    frame = FALSE
  ) +
  tm_graticules(lwd = 0.3, col = "grey70", alpha = 0.4)
```

Use the blackout mask to mask, bounding box to crop crop, and st_as_stars -> st_as_sf to vectorize the raster

```{r}
#| warning: false

# Calculate the difference in light intensity between the two dates
VIIRS_houston_diff <- VIIRS_07 - VIIRS_16


# Classify areas with light intensity drops greater than 200 as blackouts
blackout_mask <- classify(VIIRS_houston_diff,
                      matrix(c(-Inf, 200, NA,  # Not a blackout
                               200, Inf, 1),   # Blackout
                             ncol = 3, byrow = TRUE))

# Vectorize the blackout mask and fix any invalid geometries
blackout_vector <- st_as_stars(blackout_mask) %>% 
  st_as_sf() %>% 
  st_make_valid()

# Crop the vectorized blackout mask to the Houston area
blackout_houston <- st_crop(blackout_vector, HOUSTON_BBOX)

# Reproject to EPSG:3083 (NAD83 / Texas Centric Albers Equal Area)
blackout_houston <- st_transform(blackout_houston, crs = 3083)

# Verify cropping was successful
if(nrow(blackout_houston) == 0) {
  stop("No blackout areas remain after cropping to Houston")
}
```

Plot the resulting raster

```{r}
tm_basemap("CartoDB.VoyagerNoLabels") + 
  tm_shape(blackout_houston) +
  tm_polygons(fill = "red",
              col = "darkred") +
  tm_title("Initial Blackout Areas in Houston", size = 1.2) +
  tm_layout(legend.show = FALSE) +
  tm_scalebar(position = c("left", "bottom")) +
  tm_compass(position = c("right", "top"))
```

### Exclude highways from the cropped blackout mask

Highways may show reduced light intensity due to decreased traffic rather than power outages. Exclude areas within 200m of highways.

```{r}
#| warning: false

# Reproject highways to match blackout data (EPSG:3083)
highways <- st_transform(highways, crs = 3083)

# Join all the highway geometries into one and then buffer by 200m
highway_buffer <- highways %>% 
  st_union() %>% 
  st_buffer(dist = 200)

# Remove highway buffer zones from blackout areas
blackout_final <- st_difference(blackout_houston, highway_buffer)

# Verify data remains after highway exclusion
if(nrow(blackout_final) == 0) {
  warning("No blackout areas remain after excluding highways")
}
```

Map blackout area excluded highways

```{r}
tm_basemap("CartoDB.VoyagerNoLabels") +
tm_shape(highway_buffer) +
  tm_borders(col = "#895129", col_alpha = 0.5, lwd = 1) +
tm_shape(blackout_final) +
  tm_polygons(col = "red",
              border.col = "darkred",
              title = "Blackout Area (>200 nW cm⁻² sr⁻¹ drop)") +
  tm_layout(main.title = "Houston Power Outage: Blackout Areas (Highways Excluded)",
            main.title.size = 1.2,
            legend.title.size = 1,
            legend.position = c("right", "bottom")) +
  tm_scale_bar(position = c("left", "bottom")) +
  tm_compass(position = c("right", "top")) +
  tm_add_legend(type = c("lines"),
                labels = c("Highway buffer (200m)"),
                col = c("#895129")) +
  tm_add_legend(type = c("symbols"),
                labels = c("Blackout areas"),
                fill = c("red"),
                col = "darkred",
                size = 0.5)
```

### Identify the number of homes impacted by blackouts

```{r}
#| warning: false

# Reproject houses to match blackout crs (EPSG:3083) before intersecting
houses <- st_transform(houses, crs = 3083)

# Extract homes that intersect blackout areas
blackout_homes <- st_intersection(houses, blackout_final)

# Count number of impacted homes
n_blackout_homes <- length(unique(blackout_homes$osm_id))

n_total_homes <- nrow(houses)

# Calculate percent impacted
pct_impacted <- round((n_blackout_homes / n_total_homes) * 100, 1)
```


**Results:** Approximately **`r format(n_blackout_homes, big.mark = ",")`** homes (`r pct_impacted`% of residential buildings in the dataset) experienced blackouts during the February 2021 storms.

### Map impacted homes

```{r}
#| warning: false

# Create points for all homes for plotting
all_homes_points <- st_centroid(houses)

# Separate impacted and not impacted homes
not_impacted_homes <- all_homes_points[!all_homes_points$osm_id %in% blackout_homes$osm_id, ]


impacted_homes_points <- st_centroid(blackout_homes)

# Map with layers in order: blackout areas, not impacted, then impacted on top
tm_basemap("CartoDB.VoyagerNoLabels") +
  tm_shape(blackout_final) +
  tm_polygons(fill = "grey80", 
              col = "grey60") +
  tm_shape(not_impacted_homes) +
  tm_dots(fill = "#FF7F7F",
          size = 0.005,
          fill_alpha = 0.05) +
  tm_shape(impacted_homes_points) +
  tm_dots(fill = "darkred",
          size = 0.008,
          fill_alpha = 0.6) +
  tm_title("Homes in Houston Impacted by Blackouts", size = 1) +
  tm_layout(legend.outside = TRUE) +
  tm_add_legend(type = "polygons",
                labels = "Blackout area",
                fill = "grey80",
                col = "grey60",
                size = 0.2) +
  tm_add_legend(type = "symbols",
                labels = c("Impacted homes", "Not impacted homes"),
                fill = c("darkred", "#FF7F7F"),
                shape = 19,
                size = 0.5) +
  tm_scalebar(position = c("left", "bottom")) +
  tm_compass(position = c("right", "top"))
```

### Analyze census tracts
Investigate whether census tracts with different median household incomes experienced differential blackout impacts.

```{r}
#| warning: false

# Transform census data to match blackout CRS (EPSG:3083)
socio <- st_transform(socio, crs = 3083)

# Crop to Houston
houston_tracts <- st_crop(socio, st_bbox(blackout_houston))

# Count number of impacted homes per census tract
houston_tracts$n_homes <- as.numeric(lengths(st_intersects(houston_tracts, blackout_homes)))

# Create categories for visualization
houston_tracts$impact_category <- cut(houston_tracts$n_homes,
                                       breaks = c(1, 10, 50, 150, 500, 1500, Inf),
                                       labels = c("Very Low (1-10)",
                                                 "Low (11-50)", 
                                                 "Medium (51-150)", 
                                                 "High (151-500)",
                                                 "Very High (501-1,500)",
                                                 "Extreme (>1,500)"))

# Create blackout indicator
houston_tracts$blackout <- factor(houston_tracts$n_homes > 0,
                                  levels = c(TRUE, FALSE),
                                  labels = c("Blackout", "No Blackout"))

# Verify data quality
if(sum(!is.na(houston_tracts$B19013e1)) == 0) {
  stop("No valid median income data found")
}
```

```{r}
# Summary statistics
cat("Census tract summary:\n")
cat("Total tracts:", nrow(houston_tracts), "\n")
cat("Tracts with blackouts:", sum(houston_tracts$n_homes > 0), "\n")
cat("Tracts without blackouts:", sum(houston_tracts$n_homes == 0), "\n")
```

```{r}
# Map the census tracts that lost power
tm_basemap("CartoDB.VoyagerNoLabels") + 
  tm_shape(houston_tracts) +
  tm_fill(fill = "impact_category",
          fill.scale = tm_scale(values = c("#fee8c8", "#fdd49e", "#fc8d59", "#e34a33", "#b30000", "darkred"),
                                value.na = "transparent"),
          fill.legend = tm_legend(title = "Impacted Homes\nper Tract")) +
  tm_borders(col = "grey40", lwd = 0.01) +
  tm_title("Census Tracts: Number of Homes Impacted by Blackouts") +
  tm_scalebar(position = c("left", "top")) +
  tm_compass(position = c("right", "bottom")) +
  tm_layout(legend.outside = TRUE)

```

### Map of census tracts impacted by blackouts

```{r}
# Filter to only tracts with blackouts
blackout_tracts_only <- houston_tracts %>% 
  filter(blackout == "Blackout")

# Plot tracts with blackouts
tm_basemap("CartoDB.VoyagerNoLabels") +
tm_shape(blackout_tracts_only) +
  tm_fill(fill = "darkred",
          fill_alpha = 0.9) +
  tm_borders(col = "white", lwd = 0.2) +
  tm_title("Census Tracts with Blackouts in Houston") +
  tm_scalebar(position = c("left", "top")) +
  tm_compass(position = c("right", "bottom")) +
  tm_layout(legend.show = FALSE)
```
### Compare income distributions

```{r}
#| warning: false

# Remove NA from income values
houston_tracts_no_na <- houston_tracts %>% filter(!is.na(B19013e1))

# Create comparison plot
ggplot(houston_tracts_no_na, aes(x = blackout, y = B19013e1, fill = blackout)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(name = "Blackout Status", 
                    values = c("Blackout" = "#D55E00", 
                               "No Blackout" = "#999999")) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(title = "Median Household Income by Blackout Status", 
       subtitle = "Houston Metropolitan Area Census Tracts", 
       x = "Blackout Status", 
       y = "Median Household Income (USD, 2019 ACS)", 
       caption = "Source: U.S. Census Bureau ACS 5-Year Estimates (2015-2019)") +
  theme_minimal() +
  theme(legend.position = "right", 
        plot.title = element_text(size = 14, face = "bold"), 
        axis.title = element_text(size = 11))
```

### Summary statistics

```{r}
income_stats <- houston_tracts %>%
  st_drop_geometry() %>%
  filter(!is.na(B19013e1)) %>%
  group_by(blackout) %>%
  summarise(
    `Number of Tracts` = n(),
    `Median Income` = median(B19013e1),
    `Mean Income` = mean(B19013e1),
    `SD Income` = sd(B19013e1)
  )

kable(income_stats, 
      caption = "Income Statistics by Blackout Status",
      digits = 0,
      format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# Discussion and Limitations

## Key Findings

This analysis identified approximately **`r format(n_blackout_homes, big.mark = ",")`** homes in the Houston metropolitan area that likely experienced power outages during the February 2021 winter storms. By comparing night light intensity before (February 7) and after (February 16) the storms, we created a spatial mask of blackout areas and linked these to residential building locations.

The income analysis reveals that census tracts experiencing blackouts had a median household income of **`r format(income_stats$"Median Income"[income_stats$blackout == "Blackout"], big.mark = ",")`**, compared to **`r format(income_stats$"Median Income"[income_stats$blackout == "No Blackout"], big.mark = ",")`** for tracts without blackouts, suggesting ... COMPLETE THIS ANALSYS.
